Category-based and location-based volitional covert attention affect memory at different timescales

### 2023-03-26 paper/main.tex
added discussion figure, backing up progress on discussion section [WIP]

Simply by encoding their prior experiences into memory, we reasoned that
participants should rate \textit{any} presented images as more familiar than
novel images, regardless of whether they were following the attention cues. We
confirmed that this prediction held in both the sustained ($t(29) = 8.856, p <
0.001$) and variable ($t(22) = 5.144, p < 0.001$) conditions. In addition, to
the extent that participants were following the attention cues, there
\textit{internal} experiences of each image should depend on their internal
focus of attention during each image presentation. For example, we expected
that the attended-category component of the composite image at the attended
location might be better recognized than the other composite image components.
Indeed, participants in both experimental conditions rated these ``fully
attended'' images as more familiar than category-matched image components from
unattended locations (sustained: $t(29) = 6.893, p < 0.001$; variable: $t(22) =
6.938, p < 0.001$), location-matched images from the unattended category
(sustained: $t(29) = 6.710, p < 0.001$; variable: $t(22) = 7.633, p < 0.001$),
unattended images that were neither from the attended category nor the attended
location (sustained: $t(29) = 8.470, p < 0.001$; variable: $t(22) = 7.256, p <
0.001$), or novel images they had never seen before (sustained: $t(29) =
10.259, p < 0.001$; variable: $t(22) = 7.874, p < 0.001$). We also observed a
``boost'' in familiarity over novel images for attended \textit{or} unattended
images from the cued category (sustained: $t(29) = 10.578, p < 0.001$;
variable: $t(22) = 7.354, p < 0.001$) and location (sustained: $t(29) = 8.917,
p < 0.001$; variable: $t(22) = 5.726, p < 0.001$).

@@

Next, we turned to identifying potential differences in participants' behaviors
across the two experimental conditions. The main difference between the
conditions' procedures was in how long participants were asked to maintain the
same focus of category-based and location-based attention, across successive
image presentations. Therefore, differences in participants' behaviors across
the two conditions might reflect differences in the timescales of the relevant
cognitive processes. We compared participants' familiarity ratings for images
at each attention level across the two conditions. We saw no evidence that
people rated fully attended images ($t(51) = -0.649, p = 0.519$),
attended-category (but not location) images ($t(51) = 1.163, p = 0.250$),
unattended category and location images ($t(51) = -1.600, p = 0.116$), or novel
images ($t(51) = -0.435, p = 0.665$) differently across the two conditions.
However, participants in the variable attention condition rated
attended-location (but unattended category) images as more familiar than
participants in the sustained attention condition ($t(51) = 2.174, p = 0.034$).
This suggests that category-based versus location-based attention may operate
over different timescales.

@@

\textbf{JRM NOTE: STOPPED HERE}

We ran two covert attention experiments that asked participants to sustain or
vary the focus of their covert attention, respectively. When participants held
the focus of their feature-based (face versus scene) and location-based (left
versus right) attention for a sustained interval, they judged stimuli they had
seen as familiar when they overlapped with respect to the features and
locations they had attended. The increase in familiarity was larger for
attended-feature images than attended-location images. The increase also
extended to novel stimuli from the attended image category. By contrast, when
participants varied the focus of their feature-based and location-based
attention more rapidly, the boost in familiarity for feature-matched stimuli
was smaller than that for location-matched stimuli, and did not extend to novel
stimuli. Our findings suggest that participants were able to more rapidly
modulate their focus of location-based attention than their focus of
feature-based attention. The tuning of location-based attention appears to be
mediated by enhanced encoding and faster processing at the attended location.
The tuning of feature-based attention appears to be mediated by a suppression
in the encoding and processing of unattended stimulus features. This
suppression effect also affects how new stimuli are processed, and it persists
for a longer duration following an interval when the focus of feature-based
attention was held constant over a longer duration. Taken together, our
findings suggest that feature-based and location-based attention are mediated
by different mechanisms and affect memory at different timescales
(Fig.~\ref{fig:model}).

>>>>

Simply by encoding their prior experiences into memory, we reasoned that
participants should rate \textit{any} presented images as more familiar than
novel images, regardless of whether they were following the attention cues. We
confirmed that this prediction held in both the sustained ($t(29) = 8.856, p <
0.001$) and variable ($t(22) = 5.144, p < 0.001$) conditions. In addition, to
the extent that participants were following the attention cues, there
\textit{internal} experiences of each image should depend on their internal
focus of attention during each image presentation. For example, we expected
that the attended-category component of the composite image at the attended
location might be better recognized than the other composite image components.
Indeed, participants in both experimental conditions rated these ``fully
attended'' images as more familiar than category-matched image components from
unattended locations (sustained: $t(29) = 6.893, p < 0.001$; variable: $t(22) =
6.938, p < 0.001$), location-matched images from the unattended category
(sustained: $t(29) = 6.710, p < 0.001$; variable: $t(22) = 7.633, p < 0.001$),
unattended images that were neither from the attended category nor the attended
location (sustained: $t(29) = 8.470, p < 0.001$; variable: $t(22) = 7.256, p <
0.001$), or novel images they had never seen before (sustained: $t(29) =
10.259, p < 0.001$; variable: $t(22) = 7.874, p < 0.001$).

@@

Next, we turned to identifying potential differences in participants' behaviors
across the two experimental conditions. The main difference between the
conditions' procedures was in how long participants were asked to maintain the
same focus of category-based and location-based attention, across successive
image presentations. Therefore, differences in participants' behaviors across
the two conditions might reflect differences in the timescales of the relevant
cognitive processes. We compared participants' familiarity ratings for images
at each attention level across the two conditions. We saw no evidence that
people rated fully attended images ($t(51) = -0.649, p = 0.519$),
attended-category (but not location) images ($t(51) = 1.163, p = 0.250$), or
novel images ($t(51) = -0.435, p = 0.665$) differently across the two
conditions. However, participants in the variable attention condition rated
attended-location (but unattended category) images as more familiar than
participants in the sustained attention condition ($t(51) = 2.174, p = 0.034$).
We found a trending effect for unattended category and location images, whereby
participants in the variable attention condition tended to rate these images as
more familiar than participants in the sustained attention condition ($t(51) =
1.600, p = 0.116$). We also observed some within-condition differences in how
participants rated partially attended versus novel images. In the sustained
attention condition, participants rated attended-category images at the
unattended location as more familiar than novel images ($t(29) = 6.205, p <
0.001$), but participants in the variable attention condition did not show this
pattern ($t(22) = 1.042, p = 0.309$). On the other hand, whereas participants
in the sustained attention condition showed no reliable differences in
familiarity between attended-location images from the unattended category and
novel images ($t(29) = 0.165, p = 0.870$), participants in the variable
attention condition rated attended-location images as more familiar than novel
images ($t(22) = 3.026, p = 0.006$). Taken together, our analyses highlight
several familiarity differences in partially attended images from the attended
category or at the attended location across the two experimental conditions.
These differences suggest that category-based versus location-based attention
operate over different timescales.

@@

We ran a covert attention experiment with two conditions that asked
participants to sustain or vary the focus of their covert attention,
respectively. We then administered recognition memory tests that asked
participants to rate the ``familiarity'' of attended, unattended, and novel
images. In our analyses, we used eyetracking data to focus in on trials where
participants were specifically varying their focus of \textit{covert} attention
(i.e., with no change in external behavior), as opposed to simply moving their
eyes to look at the to-be-attended images. In both conditions, we found that
participants recognized images from the attended category and location more
readily than unattended or novel images. This effect was substantial and robust
(in both conditions) and also held for individual image categories. Whereas
prior work has focused primarily on \textit{overt} changes in attention (e.g.,
changes in eye movements associated with attention cues), we show that what
people \textit{covertly} attend to also affects how they remember their ongoing
experiences.  Specifically, covert attention appears to boost memory encoding
such that the focus of covert attention is recognized more readily later on.

\begin{figure*}[tp]
  \centering \includegraphics[width=0.75\textwidth]{figs/discussion_schematic}

  \caption{\textbf{How do covert location-based and category-based attention
  affect memory encoding?}\textbf{A. Hypothesized timecourses of the impact of
  location-based and category-based attention on memory encoding.} Shifting the
  focus of location-based attention increases memory encoding at the attended
  locations (blue curve). This increase can be observed after 5.5~s (the
  duration of one presentation from the variable attention condition). Shifting
  the focus of category-based attention suppresses memory encoding for the
  unattended category. However, this suppression effect occurs relatively
  slowly (longer than the duration of a single image presentation in the
  experiment). \textbf{B. Location-based attention.} Focusing covert attention
  on one \textit{location} enhances encoding of stimuli at the attended
  location (red and yellow bars), regardless of stimulus category.  (This panel
  is based on the variable attention results presented in Fig.~\ref{fig:familiarity}.)
  \textbf{C. Category-based attention.}  Focusing covert attention on one \textit{category}
  supresses encoding of stimuli from the unattended category (yellow and green bars),
  regardless of spatial location.  (This panel is based on the sustained attention
  results presented in Fig.~\ref{fig:familiarity}.)}

\label{fig:discussion}
\end{figure*}

We also found that partially attended images (e.g., from the attended category
but unattended location, or at the attended location but unattended category)
were rated as more familiar than novel images. However, these encoding benefits
differed across the two experimental conditions. In the sustained attention
condition, attended-category images from the unattended location were rated as
more familiar than novel images, but attended-location images from the
unattended category were not. Participants in the variable attention condition
showed the opposite pattern. The variable attention participants rated
attended-\textit{location} images from the unattended category as more familiar
than novel images, but they rated attended-category images from the unattended
location similarly to novel images. Because the primary difference between the
sustained and variable attention conditions was the duration of the attention
cues, our analyses of partially attended images suggest that the effects of
different aspects of attention on memory may unfold over different timescales
(Fig.~\ref{fig:discussion}). Specifically, location-based attention appears to
affect memory encoding relatively quickly, which would explain why
attended-location images from both the attended \textit{and} unattended
category receive a memory encoding benefit in the variable attention condition.
In contrast, category-based attention appears to affect memory encoding more
slowly, and it appears to operate in a ``suppressive'' manner (i.e., supressing
encoding of the unattended category, as opposed to enhancing encoding of the
attended category). This would explain why unattended-category images at the
attended location are rated as \textit{less} familiar in the sustained
attention condition.

\textbf{JRM STOPPED HERE}

When participants held the focus of their category-based (face
versus place) and location-based (left versus right) attention for a sustained
interval, they judged stimuli they had seen as familiar when they overlapped
with respect to the features and locations they had attended. The increase in
familiarity was larger for attended-feature images than attended-location
images. The increase also extended to novel stimuli from the attended image
category. By contrast, when participants varied the focus of their
feature-based and location-based attention more rapidly, the boost in
familiarity for feature-matched stimuli was smaller than that for
location-matched stimuli, and did not extend to novel stimuli. Our findings
suggest that participants were able to more rapidly modulate their focus of
location-based attention than their focus of feature-based attention. The
tuning of location-based attention appears to be mediated by enhanced encoding
and faster processing at the attended location. The tuning of feature-based
attention appears to be mediated by a suppression in the encoding and
processing of unattended stimulus features. This suppression effect also
affects how new stimuli are processed, and it persists for a longer duration
following an interval when the focus of feature-based attention was held
constant over a longer duration. Taken together, our findings suggest that
feature-based and location-based attention are mediated by different mechanisms
and affect memory at different timescales (Fig.~\ref{fig:model}).
